---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# portopt

Tools for simple portfolio optimization:

* Minimum variance portfolio for a given expected return, with no limits on allocations for individual asset classes other than that they must sum to 1. (That is, shorting and leverage are allowed.)

* Minimum variance portfolio with lower and upper bounds on allocations for asset classes. It can also accommodate more-general linear constraints on asset classes. (For example, stocks plus bonds combined must be between 40% and 60% of the portfolio.)

* Efficient frontier

`portopt` cannot handle nonlinear constraints on asset allocations, and it cannot find an optimal portfolio that satisfies more-complex objectives, such as value-at-risk criteria. For my purposes, so far, these are not necessary. `portopt` uses `solve.QP` from the package `quadprog` for minimum-variance optimization. Other solvers would be required for more-complex optimization.

I wrote `portopt` because I was looking for basic portfolio optimization tools in `R`. There are many available packages, but the package that looked like it would be most useful, `PortfolioAnalytics`, recently was removed from `CRAN`. At some point I may trade up to one of the more-sophisticated portfolio optimization packages available on `CRAN`.

## portopt data sets

The package includes several data sets with expected returns, standard deviations, and a correlation matrix for multiple asset classes. These data sets are used in the examples below. Each data set is a list with two elements:

* `ersd` - a data frame with columns class (the asset class), er (expected return), and sd (standard deviation)
* `cormat` - a correlation matrix for these assets. Thus, the number of rows and number of columns equal the number of asset classes. The row names and column names are the asset-class names.

The data sets are:

* `stalebrink`
* `rvk`
* `horizon10year2017`

The examples below use these datasets.


## Installation

Install as follows:

``` r
devtools::install_github("donboyd5/portopt")
```

## Examples

### Get basic information about a dataset
This is a basic example which shows you how to solve a common problem:

```{r basic}
library("portopt")
library("magrittr")
library("tidyverse")

stalebrink
aa.wts <- c(.3, .2, .1, .15, .1, .05) # asset-allocation weights
per(stalebrink$ersd$er, aa.wts) # portfolio expected return with these weights
psd(stalebrink$cormat, stalebrink$ersd$sd, aa.wts) # portfolio standard deviation with these weights

```


### Get the minimum variance portfolio for a given expected return
This is a basic example which shows you how to solve a common problem:

```{r portfolios}
# Create several minimum-variance portfolios

# no restrictions on asset allocation - shorting and leverage allowed
minvport(.09, stalebrink$ersd, stalebrink$cormat)$portfolio

# shorting and leverage NOT allowed:
minvport(.09, stalebrink$ersd, stalebrink$cormat, 0, 1)$portfolio

# shorting and leverage NOT allowed, 40% upper bound on real estate:
minvport(.09, stalebrink$ersd, stalebrink$cormat, 0, c(1, 1, 1, .4, 1, 1))$portfolio

```


You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```


