---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# portopt

Tools for simple portfolio optimization:

* Minimum variance portfolio for a given expected return, with no limits on allocations for individual asset classes other than that they must sum to 1. (That is, shorting and leverage are allowed.)

* Minimum variance portfolio with lower and upper bounds on allocations for asset classes. It can also accommodate more-general linear constraints on asset classes. (For example, stocks plus bonds combined must be between 40% and 60% of the portfolio.)

* Efficient frontier

`portopt` cannot handle nonlinear constraints on asset allocations, and it cannot find an optimal portfolio that satisfies more-complex objectives, such as value-at-risk criteria. For my purposes, so far, these are not necessary. `portopt` uses `solve.QP` from the package `quadprog` for minimum-variance optimization. Other solvers would be required for more-complex optimization.

I wrote `portopt` because I was looking for basic portfolio optimization tools in `R`. There are many available packages, but the package that looked like it would be most useful, `PortfolioAnalytics`, recently was removed from `CRAN` although I have installed it from github. It seems very powerful. `fPortfolio` also seems powerful. I understand the basics of how to use them when we are starting with raw data - historical data on asset returns - but I haven't figured out how to use them when we only want to use summary measures to construct an efficient frontier, by using already-available expected returns and standard deviations by asset classs, plus the correlation matrix. I am sure `PortfolioAnalytics` and `fPortfolio` can do that but I haven't figured it out. At some point I may abandon this package in favor of one of these two packages.

## portopt data sets

`portopt` includes several data sets, each with expected returns, standard deviations, and a correlation matrix for multiple asset classes. These data sets are used in the examples below. Each data set is a list with two elements:

* `ersd` - a data frame with columns class (the asset class), er (expected return), and sd (standard deviation)
* `cormat` - a correlation matrix for these assets. Thus, the number of rows and number of columns equal the number of asset classes. The row names and column names are the asset-class names.

The data sets and their respective sources are:

* `stalebrink` -- Stalebrink, O. J. “Public Pension Funds and Assumed Rates of Return: An Empirical Examination of Public Sector Defined Benefit Pension Plans.” The American Review of Public Administration 44, no. 1 (January 1, 2014): 92–111. https://doi.org/10.1177/0275074012458826.

* `rvk` -- RVK. “Asset/Liability Study: Los Angeles Fire and Police Pension System.” RVK, Inc., October 2015.

* `horizon10year2017` -- “Survey of Capital Market Assumptions: 2017 Edition.” Horizon Actuarial Services, LLC, August 2017. http://www.horizonactuarial.com/uploads/3/0/4/9/30499196/horizon_cma_survey_2017_v0822.pdf.

The associated documents are in the "docs" folder of this project, on the github site

The examples below use these datasets.


## Installation

Install as follows:

``` r
devtools::install_github("donboyd5/portopt")
```

## Examples

### Get basic information about a dataset

```{r basic, warning=FALSE, message=FALSE}
library("portopt")
library("plyr")
library("magrittr")
library("Matrix")
library("quadprog")
library("tidyverse")

stalebrink
aa.wts <- c(.25, .25, .1, .15, .1, .05) # asset-allocation weights
per(stalebrink$ersd$er, aa.wts) # portfolio expected return with these weights
psd(stalebrink$cormat, stalebrink$ersd$sd, aa.wts) # portfolio standard deviation with these weights

```


### Get the minimum variance portfolio for a given expected return

```{r portfolios}
# Create several minimum-variance portfolios

# no restrictions on asset allocation - shorting and leverage allowed
minvport(.09, stalebrink$ersd, stalebrink$cormat)$portfolio

# shorting and leverage NOT allowed:
minvport(.09, stalebrink$ersd, stalebrink$cormat, 0, 1)$portfolio

# shorting and leverage NOT allowed, 40% upper bound on real estate:
minvport(.09, stalebrink$ersd, stalebrink$cormat, 0, c(1, 1, 1, .4, 1, 1))$portfolio

```


### Check whether correlation matrix is positive definite
While it can be fixed, the differences between the original and adjusted correlation matrices can be large.
This requires investigation.
```{r check}
is.PD(stalebrink$cormat) # good

is.PD(rvk$cormat) # bad
cormat2 <- makePDcorr(rvk$cormat)
is.PD(cormat2)
# compare:
rvk$cormat %>% round(., 2)
cormat2 %>% round(., 2)
(cormat2 - rvk$cormat) %>% round(., 2)

```




### Get and graph the efficient frontier for Stalebrink assumptions

```{r efrontier}

ds <- stalebrink

ef.nobound <- efrontier(seq(.00, .30, .0025), ds$ersd, ds$cormat)
ef.noshort <- efrontier(seq(.00, .30, .0025), ds$ersd, ds$cormat, 0, 1)

# create a data frame with just the data we want
efdf <- bind_rows(ef.nobound$efrontier %>% mutate(rule="no bounds on allocation"),
                  ef.noshort$efrontier %>% mutate(rule="no shorts or leverage")) %>%
  filter(type=="high", !is.na(per))

ggplot() +
  geom_line(data=efdf, aes(psd, per, colour=rule)) +
  scale_x_continuous(name="Standard deviation", breaks=seq(0, .5, .025), limits=c(0, .27), labels = scales::percent) +
  scale_y_continuous(name="Expected return", breaks=seq(0, .5, .01), limits=c(0, .16), labels = scales::percent) +
  ggtitle("Efficient frontier for the Stalebrink capital market assumptions", subtitle="With and without bounds on asset allocations") +
  # now add the asset-class points
  geom_point(data=ef.nobound$ersd, aes(x=sd, y=er)) +
  geom_text(data=ef.nobound$ersd, aes(x=sd, y=er, label=class), nudge_y = +.003) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = .075, linetype="dashed")

```


### Repeat for the Horizon 2017 10-year assumptions
```{r}

ds <- horizon10year2017

ef.nobound <- efrontier(seq(.00, .30, .0025), ds$ersd, ds$cormat)
ef.noshort <- efrontier(seq(.00, .30, .0025), ds$ersd, ds$cormat, 0, 1)

# create a data frame with just the data we want
efdf <- bind_rows(ef.nobound$efrontier %>% mutate(rule="no bounds on allocation"),
                  ef.noshort$efrontier %>% mutate(rule="no shorts or leverage")) %>%
  filter(type=="high", !is.na(per))

ggplot() +
  geom_line(data=efdf, aes(psd, per, colour=rule)) +
  scale_x_continuous(name="Standard deviation", breaks=seq(0, .5, .025), limits=c(0, .27), labels = scales::percent) +
  scale_y_continuous(name="Expected return", breaks=seq(0, .5, .01), limits=c(0, .16), labels = scales::percent) +
  ggtitle("Efficient frontier for the Horizon 2017 average 10-year capital market assumptions", subtitle="With and without bounds on asset allocations") +
  # now add the asset-class points
  geom_point(data=ef.nobound$ersd, aes(x=sd, y=er)) +
  geom_text(data=ef.nobound$ersd, aes(x=sd, y=er, label=class), nudge_y = +.003) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = .075, linetype="dashed")

```


### Asset weights for the Stalebrink assumtions
```{r}
ds <- stalebrink

ef.nobound <- efrontier(seq(.00, .30, .0025), ds$ersd, ds$cormat)
ef.noshort <- efrontier(seq(.00, .30, .0025), ds$ersd, ds$cormat, 0, 1)

# Examine asset class weights for different classes
ef.noshort$weights %>%
  filter(class %in% c("stocks", "cash", "bonds.dom")) %>%
  ggplot(aes(er.target, asset.weight, colour=class)) +
  geom_line() +
  ggtitle("Stalebrink weights with no shorting or leverage")

# Compare weights for an asset class, optimized with and without bounds
aclass <- "stocks"
wts <- bind_rows(ef.nobound$weights %>% mutate(rule="no bounds on allocation"),
                 ef.noshort$weights %>% mutate(rule="no shorts or leverage")) %>%
       filter(class==aclass)
     
wts %>%
       ggplot(aes(er.target, asset.weight, colour=rule)) +
       geom_line() +
       geom_hline(yintercept=0) +
       scale_x_continuous(breaks=seq(0, 1, .05)) +
       ggtitle(paste0(aclass, " weights, Stalebrink assumptions"))

```



